Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex
Re-write impact

%\textcolor{red}{This comes a bit from nowhere. Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research. We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution. In the first case, the analysis use norms blabla. In the second, we use KL"}
{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
$\KL(p||q) = +\infty$ if the model $q(x)=0$ for some $x$ from the true distribution$p(x)>0$.
}
has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
Hence the natural question:
%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution. }
% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}

@@

{\bf Impact.}
While a solution to the fundamental problem~\eqref{problem} would be of independent interest -- e.g. assessing the importance of the prior in MAP with few samples-- we believe it would have repercussions to other fields. 
For instance, it could enable stochastic optimization to tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression) \citep{bach2013nonstronglyconvex}.
%Or it could help developing accelerated (stochastic?) algorithms for barrier objectives  overcoming the limitations of non-Euclidean geometry \citep{dragomir2021optimal}. \rlp{Fred I need your feeedback on that one.}

% outcomes
% quantifying the importance of the prior, and few sample behavior, for instance in Gaussians
% stochastic algorithms for barrier objectives
% such as a lot of GLM
% accelerated methods ?

@@

We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
This is true for
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.

>>>>

%\textcolor{red}{This comes a bit from nowhere. Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research. We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution. In the first case, the analysis use norms blabla. In the second, we use KL"}
{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{For $p$ and $q$ continuous densities of the ground truth and a model,
$\KL(p||q) = +\infty$ if $q(x)=0$ for some $x$ such that $p(x)>0$.
}
has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
Hence the fundamental question:
%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution. }
% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}

@@

{\bf Expected Ouctomes.}
A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime, for instance for Gaussians $\cN(m,\sigma^2)$. 
More generally it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
It could also help in assessing the impact of alternative forms of regularization (prior) for these models.

%Or it could help developing accelerated (stochastic?) algorithms for barrier objectives  overcoming the limitations of non-Euclidean geometry \citep{dragomir2021optimal}. \rlp{Fred I need your feedback on that one.}

@@

We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
This is true for
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.
This is true for categorical or Dirichlet distributions.
